### 2017-07-06 paper/conclu.tex
some changes of bibliography

\smallskip\noindent{\em Probabilistic programs and termination.}
In early works the termination for concurrent probabilistic programs was studied 
as fairness~\cite{SPH84}, which ignored precise probabilities.
For countable state space a sound and complete characterization of almost-sure termination 
was presented in~\cite{HS85}, but nondeterminism was absent.
A sound and complete method for proving termination of finite-state programs
was given in~\cite{EGK12}.
For probablistic programs with countable state space and without 
nondeterminism, the {\em Lyapunov ranking functions} provide a sound and 
complete method to prove positive termination~\cite{BG05,Foster53}.
For probabilistic programs with nondeterminism, but restricted to discrete probabilistic
choices, the termination problem was studied 
in~\cite{MM04,MM05}.
The RSM-based (ranking supermatingale-based) approach extending ranking functions was first 
presented in~\cite{SriramCAV} for probabilistic programs without non-determinism,
but with real-valued variables, and its extension for probabilistic programs
with non-determinism has been studied 
in~\cite{HolgerPOPL,CF17,CFNH16,CFG16,CNZ17,MM16:proofrule-arxiv}. 
Supermartingales were also 
considered for other liveness and safety 
properties~\cite{CVS16:martingale-recurrence-persistence,BEFH16:doob}.
While all these results deeply clarify the role of RSMs for probabilistic programs, 
the notion of lexicographic RSMs to obtain a practical approach for termination 
analysis for probabilistic programs has not been studied before, which we consider in 
this work.

>>>>

\smallskip\noindent{\em Probabilistic programs and termination.}
In early works the termination for concurrent probabilistic programs was studied 
as fairness~\cite{SPH84}, which ignored precise probabilities.
For countable state space a sound and complete characterization of almost-sure termination 
was presented in~\cite{HS85}, but nondeterminism was absent.
A sound and complete method for proving termination of finite-state programs
was given in~\cite{EGK12}.
For probablistic programs with countable state space and without 
nondeterminism, the {\em Lyapunov ranking functions} provide a sound and 
complete method to prove positive termination~\cite{BG05,Foster53}.
For probabilistic programs with nondeterminism, but restricted to discrete probabilistic
choices, the termination problem was studied 
in~\cite{MM04,MM05}.
The RSM-based (ranking supermatingale-based) approach extending ranking functions was first 
presented in~\cite{SriramCAV} for probabilistic programs without non-determinism,
but with real-valued variables, and its extension for probabilistic programs
with non-determinism has been studied 
in~\cite{HolgerPOPL,CF17,CFNH16:prob-termination,CFG16,CNZ17,MM16:proofrule-arxiv}. 
Supermartingales were also 
considered for other liveness and safety 
properties~\cite{CVS16:martingale-recurrence-persistence,BEFH16:doob}.
While all these results deeply clarify the role of RSMs for probabilistic programs, 
the notion of lexicographic RSMs to obtain a practical approach for termination 
analysis for probabilistic programs has not been studied before, which we consider in 
this work.

### 2017-07-06 paper/experiments.tex
some changes of bibliography

\section{Experimental Results}\label{sec:experiments}
We have implemented our algorithm (in C++) and present two sets of experimental results.
For all the experimental results we use the tool Aspic~\cite{FG10:aspic} for 
invariant generation, 
and our algorithm requires linear-programming solver for which we use 
CPlex~\cite{cplex}.
All our experimental results were obtained on the following platform: 
Ubuntu16.04, 7.7GB, Intel-Core i3-4130 CPU 3.40GHz QuadCore 64-bit.
%{\bf KRISH: TWO CITATIONS.}

>>>>

\section{Experimental Results}\label{sec:experiments}
We have implemented our algorithm (in C++) and present two sets of experimental results.
For all the experimental results we use the tool \textsc{Aspic}~\cite{FG10:aspic} for 
invariant generation, 
and our algorithm requires linear-programming solver for which we use 
\textsc{CPlex}~\cite{cplex}.
All our experimental results were obtained on the following platform: 
Ubuntu16.04, 7.7GB, Intel-Core i3-4130 CPU 3.40GHz QuadCore 64-bit.
%{\bf KRISH: TWO CITATIONS.}

### 2017-07-06 paper/introduction.tex
some changes of bibliography

\smallskip\noindent{\em Ranking functions and ranking supermartingales (RSMs).}
The key technique that applies for liveness analysis of non-probabilistic programs is 
the notion of {\em ranking functions}, which provides a sound and complete 
method for termination of non-probabilistic programs~\cite{rwfloyd1967programs}.
There exist a wide variety of approaches for construction 
of ranking functions for non-probabilistic programs~\cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/ColonS01,DBLP:conf/vmcai/PodelskiR04,DBLP:conf/pods/SohnG91}.
The generalization of ranking functions to probabilistic programs is achieved through the
{\em ranking supermartingales (RSMs)}~\cite{SriramCAV,HolgerPOPL,CF17}.
The ranking supermartingales provide a powerful and automated approach for termination 
analysis of probabilistic programs, and algorithmic approaches for special cases such as 
linear and polynomial RSMs have also been considered~\cite{SriramCAV,CFNH16,CFG16,CNZ17}.

\smallskip\noindent{\em Practical limitations of existing approaches.} 
While an impressive set of theoretical results related to RSMs has been
established~\cite{SriramCAV,HolgerPOPL,CF17,CFNH16,CFG16,CNZ17}, 
for probabilistic programs with nondeterminism the current approaches  
are only applicable to academic examples of variants of random walks. 
The key reason can be understood as follows: even for non-probabilistic programs while 
ranking functions are sound and complete, they do not necessarily provide a practical 
approach. This is because to prove termination, a witness in the form of a 
ranking function has to be computed: to this automatically, ranking functions 
of a restricted shape (such as linear ranking functions) have to be considered, 
and 1-dimensional ranking functions of a restricted type can only prove 
termination of a limited class of programs. 
In contrast, as a practical and scalable approach for non-probabilistic programs the 
notion of lexicographic ranking functions has been widely 
studied~\cite{CSZ13,ADFG10:lexicographic,GMR15:rank-extremal,BCIKP16:T2}. 
Algorithmic approaches for linear lexicographic ranking functions allow the 
termination analysis 
to be applicable to real-world non-probabilistic programs (after abstraction).
However both the theoretical foundations as well as practical approaches related to 
such lexicographic ranking functions are completely missing for probabilistic programs,
which we address in this work.

>>>>

\smallskip\noindent{\em Ranking functions and ranking supermartingales (RSMs).}
The key technique that applies for liveness analysis of non-probabilistic programs is 
the notion of {\em ranking functions}, which provides a sound and complete 
method for termination of non-probabilistic programs~\cite{rwfloyd1967programs}.
There exist a wide variety of approaches for construction 
of ranking functions for non-probabilistic programs~\cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/ColonS01,DBLP:conf/vmcai/PodelskiR04,DBLP:conf/pods/SohnG91}.
The generalization of ranking functions to probabilistic programs is achieved through the
{\em ranking supermartingales (RSMs)}~\cite{SriramCAV,HolgerPOPL,CF17}.
The ranking supermartingales provide a powerful and automated approach for termination 
analysis of probabilistic programs, and algorithmic approaches for special cases such as 
linear and polynomial RSMs have also been considered~\cite{SriramCAV,CFNH16:prob-termination,CFG16,CNZ17}.

\smallskip\noindent{\em Practical limitations of existing approaches.} 
While an impressive set of theoretical results related to RSMs has been
established~\cite{SriramCAV,HolgerPOPL,CF17,CFNH16:prob-termination,CFG16,CNZ17}, 
for probabilistic programs with nondeterminism the current approaches  
are only applicable to academic examples of variants of random walks. 
The key reason can be understood as follows: even for non-probabilistic programs while 
ranking functions are sound and complete, they do not necessarily provide a practical 
approach. This is because to prove termination, a witness in the form of a 
ranking function has to be computed: to this automatically, ranking functions 
of a restricted shape (such as linear ranking functions) have to be considered, 
and 1-dimensional ranking functions of a restricted type can only prove 
termination of a limited class of programs. 
In contrast, as a practical and scalable approach for non-probabilistic programs the 
notion of lexicographic ranking functions has been widely 
studied~\cite{CSZ13,ADFG10:lexicographic,GMR15:rank-extremal,BCIKP16:T2}. 
Algorithmic approaches for linear lexicographic ranking functions allow the 
termination analysis 
to be applicable to real-world non-probabilistic programs (after abstraction).
However both the theoretical foundations as well as practical approaches related to 
such lexicographic ranking functions are completely missing for probabilistic programs,
which we address in this work.
